ExpressionView - an interactive viewer for biclusters identified in gene expression data

### 2010-01-11 expressionviewnote/expressionview.tex
Small corrections in the ExpressionView application note.

\subsection{Gene expression data and biclusters}
ExpressionView is designed to work with gene expression data in the form of a Bioconductor ExpressionSet. This class provides a user-friendly way to access the actual gene expression matrix and its associated metadata. There is a variety of biclustering algorithms described in the literature~\citep{madeira04,prelic06}, several of which have been implemented as R packages. ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03,csardi09} and the methods available in the Biclust package~\citep{kaiser08}. Since the structure of biclustering results is independent of the algorithm, an extension to other methods is straightforward.

@@

Since this reordering is an interesting problem on its own, which to the best of our knowledge has not been studied in the literature, we briefly outline our strategy here. Noting that the rows and the columns can be ordered separately, the problem reduces to finding the optimal arrangement of a set of $n$ elements that are part of at least one of $m$ clusters. The quality of the order is defined as the sum over the maximal number of neighbouring elements in every cluster and the tasks thus consists in maximizing this quantity by changing the order of the elements. Starting from an initial configuration determined by similarity, this is achieved by either shifting well aligned subsequences of a given cluster, herby enlarging the longest contiguous part, or permuting the individual elements of a given subcluster.

@@

\paragraph{Funding\textcolon} This work was supported by ...

@@

\bibitem[Cs{\'a}rdi et~al.(2009)]{csardi09}
Cs{\'a}rdi,G., Kutalik,Z. and Bergmann,S. (2009)
Modular analysis of gene expression data with R.
\href{http://dx.doi.org/}{\emph{Bioinformatics}, ...}

>>>>

\subsection{Gene expression data and biclusters}
ExpressionView is designed to work with gene expression data in the form of a Bioconductor ExpressionSet. This class provides a user-friendly way to access the actual gene expression matrix and its associated metadata. There is a variety of biclustering algorithms described in the literature~\citep{madeira04,prelic06}, several of which have been implemented as R packages. ExpressionView can treat results obtained by the iterative signature algorithm (ISA)~\citep{bergmann03}%,csardi09
and the methods available in the Biclust package~\citep{kaiser08}. Since the structure of biclustering results is independent of the algorithm, an extension to other methods is straightforward.

@@

Since this reordering is an interesting problem on its own, which to the best of our knowledge has not been studied in the literature, we briefly outline our strategy here. Noting that the rows and the columns can be ordered separately, the problem reduces to finding the optimal arrangement of a set of $n$ elements that are part of at least one of $m$ clusters. The quality of the order is defined as the sum over the maximal number of neighbouring elements in every cluster and the tasks thus consists in maximizing this quantity by changing the order of the elements. Starting from an initial configuration determined by similarity, this is achieved by either shifting well aligned subsequences of a given cluster, hereby enlarging the longest contiguous part, or permuting the individual elements of a given subcluster.

@@

\paragraph{Funding\textcolon} The authors are grateful to the Swiss
Institute of Bioinformatics, the Swiss National Science Foundation
(3100AO-116323/1), and the European Framework Project 6 (through
the EuroDia and AnEuploidy projects).

@@

%\bibitem[Cs{\'a}rdi et~al.(2009)]{csardi09}
%Cs{\'a}rdi,G., Kutalik,Z. and Bergmann,S. (2009)
%Modular analysis of gene expression data with R.
%\href{http://dx.doi.org/}{\emph{Bioinformatics}, ...}
